---
title: "Predicting Disordered Regions using Folding Confidence Values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{disprot_vs_alphafold}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(disfold)
library(dplyr)
library(knitr)
library(bio3d)
library(ggplot2)
library(here)
data(disprot23)
PROJECT_ROOT=here()
scratchdir=file.path(PROJECT_ROOT, 'data-raw/afdb')
URL_BASE = "https://alphafold.ebi.ac.uk/files/"
dir.create(scratchdir, recursive = TRUE) # create scratch directory
```

## Technicalities

The value of the `here()` function is: `r here()`. 
Created scratch directory for download of protein structure data files: `r scratchdir`.

## Overview

```{r}
disprot23 %>% head() %>% kable()
```


## Accession Codes

```{r}
accessions = sort(unique(disprot23$acc))
accessions = head(accessions, 100)
```

The Disprot dataset (2023-12) consists of `r nrow(disprot23)` annotated regions corresponding to `r length(accessions)` unique UniProt ids.

# Loop over protein chains

```{r}
# typical URL for accession P00520: https://alphafold.ebi.ac.uk/files/AF-P00520-F1-model_v4.cif


regions_to_residues <- function(starts, ends) {
    stopifnot(length(starts) == length(ends))
    n = length(starts)
    result = c()
    for (i in 1:n) {
        result = c(result, starts[i]:ends[i])
    }
    result
}


# accessions = c('P10902')
ca_atoms_all = NULL
    
for (acc in accessions) {
    fname = paste0("AF-",acc, "-F1-model_v4.pdb")
    url = paste0(URL_BASE, fname)
    destfile = file.path(scratchdir, fname)
    if (! file.exists(destfile)) {
        cat("Downloading", acc, 'from', url, '...')
        try(download.file(url, destfile), silent=TRUE)
    }
    if (!file.exists(destfile)) {
        next
    }
    pdb <- read.pdb(destfile) # alternative: read.cif
    ca_atoms <- pdb$atom[pdb$atom$elety == "CA", ]
    # print(head(ca_atoms))
    disregions = disprot23[disprot23$acc==acc, ]
    # print(disregions)
    disresidues = regions_to_residues(disregions$start, disregions$end)
    dis_confidences = ca_atoms[disresidues, 'b']
    ca_atoms[['structured']] = 1
    ca_atoms[disresidues, 'structured'] = 0
    ca_atoms[['acc']] = acc
    if (is.null(ca_atoms_all)) {
        ca_atoms_all = ca_atoms
    } else {
     ca_atoms_all = rbind(ca_atoms_all, ca_atoms)
    }
}

```

## Visualizing Results

```{r}
ca_atoms_all %>% ggplot(aes(x=factor(structured), y=b)) + geom_violin()
```
## Correlation

The correlation between The AF2 confidence and the disordered region annotation is `r round(cor(ca_atoms_all[['structured']], ca_atoms_all[['b']]), 4)` or $r^2=$ `r r=cor(ca_atoms_all[['structured']], ca_atoms_all[['b']]); round(r*r,4)`


## Prediction based on Simple Threshold

```{r}
struct_threshold=75
ca_atoms_all[['predstruct']] = as.numeric(ca_atoms_all[['b']] > struct_threshold)
table(ca_atoms_all[['predstruct']], ca_atoms_all[['structured']])
```
