---
title: "Predicting Disordered Regions using Folding Confidence Values"
author: "Course BIFX553-Spring2024"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{disprot_vs_alphafold}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Synopsis

This R package is focused on the analysis of disordered regions in proteins. In this R-markdown file we want to determine to what extent per-residue structure prediction confidence values from AlphaFold are associated with the two
classes of structured versus disordered regions. 

The underlying dataset contains annotations of disordered regions in terms of Uniprot protein accession numbers. It was downloaded from disprot (<https://disprot.org>) [1]. 

This markdown file is just a "writing prompt" in the sense that it helps to get some initial data loaded combined with some basic visualization. But much more could be done!

[1] DisProt in 2024: improving function annotation of intrinsically disordered proteins
Aspromonte MC, Nugnes MV, Quaglia F, Bouharoua A, DisProt Consortium, Tosatto SCE and Piovesan D (2023) Nucleic Acids Research, Database Issue


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(disfold)
library(dplyr)
library(knitr)
library(bio3d)
library(ggplot2)
library(here)
data(disprot23)
# Parameters
PROJECT_ROOT=here()
scratchdir=file.path(PROJECT_ROOT, 'data-raw/afdb')
URL_BASE = "https://alphafold.ebi.ac.uk/files/"
PROTEIN_DOWNLOAD_LIMIT = 100 # initial limit before "production" runs
# Create scratch directory for protein structures that are downloaded
dir.create(scratchdir, recursive = TRUE) # create scratch directory
```

## Technicalities

The value of the `here()` function is: `r here()`. 
Created scratch directory for download of protein structure data files: `r scratchdir`.

## Overview

```{r}
disprot23 %>% head(n=20) %>% kable()
```


## Accession Codes

```{r}
accessions = sort(unique(disprot23$acc))
accessions = head(accessions, PROTEIN_DOWNLOAD_LIMIT)
```

The Disprot dataset (2023-12) consists of `r nrow(disprot23)` annotated regions corresponding to `r length(accessions)` unique UniProt ids.

# Downloading And Analyzing Predicted Structures

```{r}
# typical URL for accession P00520: https://alphafold.ebi.ac.uk/files/AF-P00520-F1-model_v4.cif


regions_to_residues <- function(starts, ends) {
    stopifnot(length(starts) == length(ends))
    n = length(starts)
    result = c()
    for (i in 1:n) {
        result = c(result, starts[i]:ends[i])
    }
    result
}


# accessions = c('P10902')
ca_atoms_all = NULL
    
for (acc in accessions) {
    fname = paste0("AF-",acc, "-F1-model_v4.pdb")
    url = paste0(URL_BASE, fname)
    destfile = file.path(scratchdir, fname)
    if (! file.exists(destfile)) {
        cat("Downloading", acc, 'from', url, '...')
        try(download.file(url, destfile), silent=TRUE)
    }
    if (!file.exists(destfile)) {
        next
    }
    pdb <- read.pdb(destfile) # alternative: read.cif
    ca_atoms <- pdb$atom[pdb$atom$elety == "CA", ]
    if (!is.data.frame(ca_atoms) || nrow(ca_atoms) < 10) {
        cat("Skipping", acc, 'due to insufficient number of C-alpha atom coordinates\n')
        next
    }
    # print(head(ca_atoms))
    disregions = disprot23[disprot23$acc==acc, ]
    # print(disregions)
    disresidues = regions_to_residues(disregions$start, disregions$end)
    dis_confidences = ca_atoms[disresidues, 'b']
    ca_atoms[['structured']] = 1
    ca_atoms[disresidues, 'structured'] = 0
    ca_atoms[['acc']] = acc
    if (is.null(ca_atoms_all)) {
        ca_atoms_all = ca_atoms
    } else {
     ca_atoms_all = rbind(ca_atoms_all, ca_atoms)
    }
}

```

## Structure Prediction Confidence vs Experimental Evidence of Protein Structure

```{r}
ca_atoms_all %>% ggplot(aes(x=factor(structured, labels=c("Disordered", "Structured")), y=b)) + geom_violin()  + xlab("Local Protein Structure") + ylab("AlphaFold Confidence Value")
```


